A Computational Account of Default-Mode Function\\by Control Theory and Reinforcement Learning

### 2016-08-04 paper.tex
some friston critics

The goal of the agent is then to optimize over the generative model $G$ (i.e over the generative density $p_G(.|s)$, so as to minimize their surprise. It turns out that a direct attempt to attack this optimization problem by gradient descent on the free-energy $F_G(s)$ is furtile: the parameter update steps are not ``very clean'', and require rather cumbersome and heavy computations. A workaround is then to introduce a second density $p_R(.|\zeta)$ called a \textit{recognition} density to work in tandem with the generative density $p_G(.|\zeta)$: the former dreams / fantacizes whilst the latter tries to generate sensations which match these dreams! This primal-dual idea, first proposed in Hinton et al. 1995, is at the heart ot the general free-energy principle that we'll introduce shortly.

@@

\subsubsection{Helmholtz machines}
...
\subsection{The wake-sleep algorithm}
...
\subsection{Friston's active-inference: an application of the Dayan's wake-sleep algorithm for training a Helmholtz machine model for the brain}
...

%% \subsection{A thermodynamic model for bounded rationality, aka robust optimality}
%% \begin{itemize}
%%   \item Recall that an agent is said to have \textit{bounded rationality} if they must take into account the cost of finding solutions to problems, and not just the utility of the final state.
%%   \item For example, consider an agent that must operate under a limited lifetime and/or computation cost.
%%   \item This is in contrast to agents with \textit{unbounded rationality} considered in classical game theory.
%%     \end{itemize}
%% The material presented is a revisit of \cite{braun2011path}. See also \cite{ortega2013thermodynamics}
%% \paragraph{Utility functions and conjugate pairs.}
%%   Let $(\Omega, \mathcal F, P)$ be a probablity space. A function $U: \mathcal F \rightarrow \mathbb R$ is said to be a \textit{utility function} for this space if the conditional utility $U(A|B) := U(A \cap B) - U(B)$ has the following propertites:
%%   \begin{itemize}
%%   \item additivity: $U(A_1 \cap A_2 | B) = U(A_1|B) + U(A_2|B)$, for all events $A_1, A_2, B \in \mathcal F$.    
%%   \item statistic: there exists a function $f_{U} :\mathbb R_+ \rightarrow \mathbb R$ such that $U(A|B) = f(P(A|B))$, for all events
%%     $A,B \in \mathcal F$.
%%   \item monotonicity: $f_{U}$ is strictly increasing.
%%     \end{itemize}

%% \begin{theorem}
%%   The only functions $f: \mathbb R_+ \rightarrow \mathbb R$ which is such that $U(A|B) \equiv f(P(A|B))$ any probability space $(\Omega, \mathcal F, P)$ and utility function $U$ thereupon are of the form
%%   \begin{equation}
%%     f = \alpha \log(.),
%%     \label{eq:boltzmann}
%%   \end{equation}
%%   where $\alpha > 0$.
%% \end{theorem}

%% \textbf{XXX: equation \eqref{eq:boltzmann} above looks like Boltzmann's formula (on his gravestone...)!}

%% Such $U$ and $P$ are said to form a \textit{conjugate pair} at temperature $\alpha$.

%% \paragraph{Example.}
%% Given a utility function $U$ on a probability space $(\Omega, \mathcal F, *)$, the \textit{Gibbs measure} at temperature $\alpha > 0$ and energy levels $(-U(\omega))_{\omega \in \Omega}$ is defined to be the probability measure (on thesame measurable space)
%%   \begin{equation}
%%     P(\omega) = \frac{1}{Z_{U}(\alpha)}\exp\left(\frac{1}{\alpha}U(\omega)\right), \; \forall \omega \in \Omega,
%%   \end{equation}
%%   where

%%   \begin{equation}
%%     Z_{U}(\alpha) := \sum_{\omega \in \Omega}\exp\left(\frac{1}{\alpha}U(\omega)\right)
%%   \end{equation}
%%   is a normalization constant called the \textit{partition function} of $U$.
%%   It's not hard to see that $U$ and the $P$ above form a conjugate pair.

%%   \paragraph{Free-utility functional.}
%%     Let $(U, P)$ be a conjugate pair at temperature $\alpha > 0$  on a measurable space $(\Omega, \mathcal F)$. Given another probability measure $P'$ on the same space, define it's \textit{free utility} as
%%     \begin{equation}
%%       J(P'|U, P) = \langle U \rangle_{P'} + \alpha \mathcal H(P'),
%%     \label{eq:free_u}
%%     \end{equation}
%%     where
%%     \begin{equation}
%%       \mathcal H(P') := \langle \log(P') \rangle_{P'} := -\sum_{\omega \in \Omega}P'(\omega)\log(P'(\omega))
%%     \end{equation}
%%     is the \textit{entropy} of $P'$ (measured in the Naperian base $e \approx 2.73$). It's not difficult to establish the upper bound
%%   \begin{equation}
%%     J(P'|U) \le J(P|U) = \sum_{\omega \in \Omega}U(\omega) =: U(\Omega).
%%   \end{equation}
%%   In particular, if $P$ is the Gibbs measure at temperature $\alpha$ corresponding to $U$, then the upper bound above reduces to the \textit{log-partition function}
%%   \begin{equation}
%%     J(P'|U) \le U(\Omega) = -\alpha \log(Z_{U}(\alpha)).
%%     \end{equation}

%%   \paragraph{The free-energy / utility principle (of Friston ?).}
%%   We are now in shape to introduce the notion of free-energy for model transitions, and a variational principle for optimizing it. Consider thus an initial system described by a conjugate pair $(U_{\text{ini}}, P_{\text{ini}})$ at temperature $\alpha > 0$. We want to transform this to a new model by adding constraints represented by the utility function $\Delta U$.  The resulting system has final utility $U_{\text{fin}} = U_{\text{ini}} + \Delta U$. The difference in free-utility is then
%%   \begin{equation}
%%     \Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}} := J_{\text{fin}} - J_{\text{ini}} = \underbrace{\langle \Delta U\rangle_{P_{\text{fin}}}}_{\textbf{accuracy}} -  \underbrace{\alpha D_{\text{KL}}(P_{\text{ini}}\|P_{\text{fin}})}_{\textbf{complexity}},
%%     \label{eq:free}
%%   \end{equation}
%%   where
%%   \begin{equation}
%%     D_{\text{KL}}(P_{\text{fin}}\|P_{\text{ini}}) := \langle \log(\P_{\text{fin}}/\P_{\text{ini}})\rangle_{\P_{\text{fin}}} := \sum_{\omega \in \Omega}\P_{\text{fin}}(\omega)\log(\P_{\text{fin}}(\omega)/\P_{\text{ini}}(\omega))
%%   \end{equation}
%%   is the Kullback-Leibler divergence, and represents the information cost (measured in energy units) of changing the initial system.
%% In the above formula, we've extensively used the fact that $(U_{\text{ini}}, P_{\text{ini}})$ is a congugate pair and so $U_{\text{ini}}(\omega) \equiv \alpha \log(P_{\text{ini}}(\omega))$ by virtue of \eqref{eq:boltzmann}.
%%   The two terms in \eqref{eq:free} (accuracy or expected gain in utility, and the complexity of the transition) can be viewed as dertiminants of bounded rational decision-making. They formalize
%% a trade-off between an expected utility $\Delta U$ (first term) and
%% the information cost of transforming Pi
%% into $P_{\text{fin}}$ (second
%% term). In this interpretation $P_{\text{ini}}$ represents an initial choice
%% probability or policy, which includes the special case of the
%% uniform distribution where the decision-maker has initially no
%% preferences between the different choices. The probability measure $P_{\text{fin}}$ is the final choice probability that we are looking for since it
%% considers the utility constraint U∗
%% that we want to optimize. We can then formulate a variational principle for bounded rationality in the probabilities $P_{\text{fin}}(\omega)$
%% \begin{equation}
%% P^*_{\text{fin}} := \underset{P_{\text{fin}}}{\text{argmax }} \Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}}
%%   \end{equation}

%% By differentiating the RHS of \eqref{eq:free} w.r.t $P_{\text{fin}}$ and setting to zero, we obtain the closed-form solution

%% \begin{equation}
%%   P^*_{\text{fin}}(\omega) \propto P_{\text{ini}}(\omega)\exp\left(\frac{1}{\alpha}\Delta U(\omega)\right).
%% \end{equation}

%% Two limit cases are worth considering.
%% \begin{itemize}
%% \item \textbf{Low-temperature regime $\alpha \approx 0$:} Here $\Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}} \approx \langle \Delta U\rangle_{P'_{\text{fin}}}$, and so it's optimal to take
%%   $$P^*_{\text{fin}}(\omega) \equiv \text{dirac}(\omega - \omega^*) = \begin{cases}1, &\mbox{if }\omega = \omega^*,\\0, &\mbox{ otherwise,}\end{cases}$$
%%   where $\omega^* := \argmax_{\omega \in \Omega}U(\omega)$. This corresponds to unbounded rational decision-making, in which the cost of transition / problem-solving is completely disregarded.
%%   \item \textbf{High-temperature regime $\alpha \rightarrow +\infty$:} In this limiting case, it's optimal to take $P^*_{\text{fin}}(\omega) \equiv P_{\text{ini}}(\omega)$, i.e the change is so costly that it's optimal to maintain the current choice probabilities.
%% \end{itemize}

%% To conclude this section, let's note that \cite{braun2011path} show how their free-energy framework (on paths) links with the well-known Hamilton-Jacobi-Bellman optimal control framework. For example, one can re-derive the Linear Quadratic Gaussian (LQG) controller, which is a generalization of the LQR in \eqref{eq:hj}...

>>>>

The goal of the brain is then to optimize over the generative model $G$ (to iteratively or analytically modify the generative density $p_G(.|s)$, so as to minimize their surprise. It turns out that a direct attempt to attack this optimization problem by gradient descent on the free-energy $F_G(s)$ is furtile: the parameter update steps are not ``very clean'', and require rather cumbersome and heavy computations. A workaround is then to introduce a second density $p_R(.|\zeta)$ called a \textit{recognition} density to work in tandem with the generative density $p_G(.|\zeta)$: the former dreams / fantacizes whilst the latter tries to generate sensations which match these dreams! This primal-dual idea, first proposed in Hinton et al. 1995, is at the heart ot the general free-energy principle that we'll introduce shortly.

@@

\subsubsection{Helmholtz machines}
% \begin{mdframed}
  \textbf{Assumption:} In both generative and recognition components of the network, there is conditional independence of  neurons in the same layer, given the data. Precisely
$$
p_G(\psi^{(l)}|\zeta) = \Pi_{k=1}^{h_l}p_G(\psi_k^{(l)} | \zeta),\hspace{1cm}
p_R(\psi^{(l)}|\zeta) = \Pi_{k=1}^{h_l}p_R(\psi_k^{(l)} |\zeta)
$$
% \end{mdframed}
\subsection{The wake-sleep algorithm}
...
\subsection{Friston's active-inference}
This is nothing but an application of the Dayan's wake-sleep algorithm for training a Helmholtz machine model of the brain...

The following critics can be made:
\begin{itemize}
  \item 
    As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
    \end{itemize}

\subsection{A thermodynamic model for bounded rationality, aka robust optimality}
\begin{itemize}
  \item Recall that an agent is said to have \textit{bounded rationality} if they must take into account the cost of finding solutions to problems, and not just the utility of the final state.
  \item For example, consider an agent that must operate under a limited lifetime and/or computation cost.
  \item This is in contrast to agents with \textit{unbounded rationality} considered in classical game theory.
    \end{itemize}
The material presented is a revisit of \cite{braun2011path}. See also \cite{ortega2013thermodynamics}
\paragraph{Utility functions and conjugate pairs.}
  Let $(\Omega, \mathcal F, P)$ be a probablity space. A function $U: \mathcal F \rightarrow \mathbb R$ is said to be a \textit{utility function} for this space if the conditional utility $U(A|B) := U(A \cap B) - U(B)$ has the following propertites:
  \begin{itemize}
  \item additivity: $U(A_1 \cap A_2 | B) = U(A_1|B) + U(A_2|B)$, for all events $A_1, A_2, B \in \mathcal F$.    
  \item statistic: there exists a function $f_{U} :\mathbb R_+ \rightarrow \mathbb R$ such that $U(A|B) = f(P(A|B))$, for all events
    $A,B \in \mathcal F$.
  \item monotonicity: $f_{U}$ is strictly increasing.
    \end{itemize}

\begin{theorem}
  The only functions $f: \mathbb R_+ \rightarrow \mathbb R$ which is such that $U(A|B) \equiv f(P(A|B))$ any probability space $(\Omega, \mathcal F, P)$ and utility function $U$ thereupon are of the form
  \begin{equation}
    f = \alpha \log(.),
    \label{eq:boltzmann}
  \end{equation}
  where $\alpha > 0$.
\end{theorem}

\textbf{XXX: equation \eqref{eq:boltzmann} above looks like Boltzmann's formula (on his gravestone...)!}

Such $U$ and $P$ are said to form a \textit{conjugate pair} at temperature $\alpha$.

\paragraph{Example.}
Given a utility function $U$ on a probability space $(\Omega, \mathcal F, *)$, the \textit{Gibbs measure} at temperature $\alpha > 0$ and energy levels $(-U(\omega))_{\omega \in \Omega}$ is defined to be the probability measure (on thesame measurable space)
  \begin{equation}
    P(\omega) = \frac{1}{Z_{U}(\alpha)}\exp\left(\frac{1}{\alpha}U(\omega)\right), \; \forall \omega \in \Omega,
  \end{equation}
  where

  \begin{equation}
    Z_{U}(\alpha) := \sum_{\omega \in \Omega}\exp\left(\frac{1}{\alpha}U(\omega)\right)
  \end{equation}
  is a normalization constant called the \textit{partition function} of $U$.
  It's not hard to see that $U$ and the $P$ above form a conjugate pair.

  \paragraph{Free-utility functional.}
    Let $(U, P)$ be a conjugate pair at temperature $\alpha > 0$  on a measurable space $(\Omega, \mathcal F)$. Given another probability measure $P'$ on the same space, define it's \textit{free utility} as
    \begin{equation}
      J(P'|U, P) = \langle U \rangle_{P'} + \alpha \mathcal H(P'),
    \label{eq:free_u}
    \end{equation}
    where
    \begin{equation}
      \mathcal H(P') := \langle \log(P') \rangle_{P'} := -\sum_{\omega \in \Omega}P'(\omega)\log(P'(\omega))
    \end{equation}
    is the \textit{entropy} of $P'$ (measured in the Naperian base $e \approx 2.73$). It's not difficult to establish the upper bound
  \begin{equation}
    J(P'|U) \le J(P|U) = \sum_{\omega \in \Omega}U(\omega) =: U(\Omega).
  \end{equation}
  In particular, if $P$ is the Gibbs measure at temperature $\alpha$ corresponding to $U$, then the upper bound above reduces to the \textit{log-partition function}
  \begin{equation}
    J(P'|U) \le U(\Omega) = -\alpha \log(Z_{U}(\alpha)).
    \end{equation}

  \paragraph{The free-energy / utility principle (of Friston ?).}
  We are now in shape to introduce the notion of free-energy for model transitions, and a variational principle for optimizing it. Consider thus an initial system described by a conjugate pair $(U_{\text{ini}}, P_{\text{ini}})$ at temperature $\alpha > 0$. We want to transform this to a new model by adding constraints represented by the utility function $\Delta U$.  The resulting system has final utility $U_{\text{fin}} = U_{\text{ini}} + \Delta U$. The difference in free-utility is then
  \begin{equation}
    \Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}} := J_{\text{fin}} - J_{\text{ini}} = \underbrace{\langle \Delta U\rangle_{P_{\text{fin}}}}_{\textbf{accuracy}} -  \underbrace{\alpha D_{\text{KL}}(P_{\text{ini}}\|P_{\text{fin}})}_{\textbf{complexity}},
    \label{eq:free}
  \end{equation}
  where
  \begin{equation}
    D_{\text{KL}}(P_{\text{fin}}\|P_{\text{ini}}) := \langle \log(\P_{\text{fin}}/\P_{\text{ini}})\rangle_{\P_{\text{fin}}} := \sum_{\omega \in \Omega}\P_{\text{fin}}(\omega)\log(\P_{\text{fin}}(\omega)/\P_{\text{ini}}(\omega))
  \end{equation}
  is the Kullback-Leibler divergence, and represents the information cost (measured in energy units) of changing the initial system.
In the above formula, we've extensively used the fact that $(U_{\text{ini}}, P_{\text{ini}})$ is a congugate pair and so $U_{\text{ini}}(\omega) \equiv \alpha \log(P_{\text{ini}}(\omega))$ by virtue of \eqref{eq:boltzmann}.
  The two terms in \eqref{eq:free} (accuracy or expected gain in utility, and the complexity of the transition) can be viewed as dertiminants of bounded rational decision-making. They formalize
a trade-off between an expected utility $\Delta U$ (first term) and
the information cost of transforming Pi
into $P_{\text{fin}}$ (second
term). In this interpretation $P_{\text{ini}}$ represents an initial choice
probability or policy, which includes the special case of the
uniform distribution where the decision-maker has initially no
preferences between the different choices. The probability measure $P_{\text{fin}}$ is the final choice probability that we are looking for since it
considers the utility constraint U∗
that we want to optimize. We can then formulate a variational principle for bounded rationality in the probabilities $P_{\text{fin}}(\omega)$
\begin{equation}
P^*_{\text{fin}} := \underset{P_{\text{fin}}}{\text{argmax }} \Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}}
  \end{equation}

By differentiating the RHS of \eqref{eq:free} w.r.t $P_{\text{fin}}$ and setting to zero, we obtain the closed-form solution

\begin{equation}
  P^*_{\text{fin}}(\omega) \propto P_{\text{ini}}(\omega)\exp\left(\frac{1}{\alpha}\Delta U(\omega)\right).
\end{equation}

Two limit cases are worth considering.
\begin{itemize}
\item \textbf{Low-temperature regime $\alpha \approx 0$:} Here $\Delta J_{(U_{\text{ini}}, P_{\text{ini}}) \rightarrow P_{\text{fin}}} \approx \langle \Delta U\rangle_{P'_{\text{fin}}}$, and so it's optimal to take
  $$P^*_{\text{fin}}(\omega) \equiv \text{dirac}(\omega - \omega^*) = \begin{cases}1, &\mbox{if }\omega = \omega^*,\\0, &\mbox{ otherwise,}\end{cases}$$
  where $\omega^* := \argmax_{\omega \in \Omega}U(\omega)$. This corresponds to unbounded rational decision-making, in which the cost of transition / problem-solving is completely disregarded.
  \item \textbf{High-temperature regime $\alpha \rightarrow +\infty$:} In this limiting case, it's optimal to take $P^*_{\text{fin}}(\omega) \equiv P_{\text{ini}}(\omega)$, i.e the change is so costly that it's optimal to maintain the current choice probabilities.
\end{itemize}

To conclude this section, let's note that \cite{braun2011path} show how their free-energy framework (on paths) links with the well-known Hamilton-Jacobi-Bellman optimal control framework. For example, one can re-derive the Linear Quadratic Gaussian (LQG) controller, which is a generalization of the LQR in \eqref{eq:hj}...
